Marlowe: implementing and analysing financial contracts on blockchain

### 2019-12-11 papers/marlowe2019/paper.tex
Shorten section 5

We can also use proof assistants to demonstrate that the Marlowe semantics presents certain desirable properties, such as that money is preserved and anything unspent is returned to users by the end of the execution of any contract. In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.

Currently, we have translated the Haskell Marlowe semantics to Isabelle as closely as possible. The main differences between the Isabelle and the Haskell semantics are that:
\begin{itemize}
    \item The Isabelle semantics uses \emph{integers for identifiers} and the Haskell semantics uses strings. We do this because handling integers in Isabelle is easier and we only compare them for equality and order (their value is irrelevant).
    \item We use a \emph{custom implementation of maps and sets} that uses lists of tuples and lists of elements respectively. We do this because Isabelle already provides many theorems that are proved for lists, so it considerably facilitates the automation of certain proofs.
\end{itemize}

@@

As part of the formalisation process, Isabelle requires functions to be terminating. While  Isabelle proves termination automatically for most functions in the semantics, that is not the case for \texttt{reductionLoop}. The reason why Isabelle cannot infer termination for this function is that it repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination  requires a proof that \texttt{reduceContractStep} will eventually do that.

The way we proved that was by establishing a measure for the size of a pair of \texttt{Contract} and \texttt{State}. We defined such a measure as follows:

@@

The reason that we need the number of accounts in the \texttt{State} as part of the measure is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}, so its \texttt{size} does not decrease.

The reason that we also multiply the size of the \texttt{Contract} by two is that the primitive \texttt{Deposit} may increase the number of accounts in the \texttt{State} by one, but it will also reduce the size of \texttt{Contract} by at least one. By multiplying the size of the \texttt{Contract} by two, the decrease in the \texttt{size} of the \texttt{Contract} compensates the increase in the \texttt{length} of the list of accounts and, thus, it ensures that the overall measure \texttt{evalBound} will decrease even in that scenario.

@@

Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of \texttt{State} that make no sense as maps:

\begin{enumerate}
    \item The lists of tuples should represen maps, so it does not make sense for two elements of the list to be tuples with the same first element (repeated keys).
    \item We want to represent each map in a unique way as a list, and so we require the tuples in the list to be sorted in ascending order by their first element. This allows us to use standard equality for comparing maps.
    \item We only expect accounts to store positive amounts of money, so the second element of each tuple in the list must be a positive integer. Whenever an account becomes empty we  delete the corresponding pair  from the list.
\end{enumerate}
\noindent
If a value of \texttt{State} satisfies the first two properties above we call it a \texttt{validState}; we also formalised the third property above as follows:

\begin{verbatim}
fun positiveMoneyInAccountOrNoAccount
             :: "AccountId ⇒ Token ⇒ Accounts ⇒ bool" where
"positiveMoneyInAccountOrNoAccount accId tok mlist =
    (case MList.lookup (accId, tok) mlist of
       None ⇒ True
     | Some x ⇒ x > 0)"
\end{verbatim}
\noindent
For a state to have positive accounts, the above function must evaluate to \texttt{True} for any \texttt{AccountId} and \texttt{Token}.

We have proved that if all the three properties are true for a given input \texttt{State}, then the properties will also be true for the resulting output \texttt{State}, when processing any \texttt{Transaction} on any \texttt{Contract}.

@@

There is a class of combinations of \texttt{State} and \texttt{Contract} that will not evolve when passed to the semantics unless a matching \texttt{Input} is passed as well. We call these combinations of \texttt{Contract} and \texttt{State} \textit{quiescent}, and we define them as follows:

\begin{verbatim}
fun isQuiescent :: "Contract ⇒ State ⇒ bool" where
"isQuiescent Close state = (accounts state = [])" |
"isQuiescent (When _ _ _) _ = True" |
"isQuiescent _ _ = False"
\end{verbatim}
\noindent
We have proved that, if an input \texttt{State} is valid and all accounts are positive, then the combination of \texttt{State} and \texttt{Contract} resulting from processing any transaction will always be quiescent.

@@

One of the dangers of using smart contracts is that a badly written one can potentially lock its funds forever. We want to avoid that feature in Marlowe altogether and, with that aim, we proved that Marlowe account system does not create or destroy any money. By the end of the contract, all the money paid to the contract must be distributed back, in some way, to a subset of the participants of the contract.
To ensure this is the case we proved two properties:

\subsubsection{Money preservation}
Money is not created or destroyed by the semantics. We state this property as follows:

\begin{verbatim}
lemma computeTransaction_preserves_money :
"validAndPositive_state state
  ⟹ moneyInState state + moneyInInputs (inputs tra)
       = moneyInTransactionOutput state (inputs tra)
               (computeTransaction tra state contract)"
\end{verbatim}
\noindent
The only consideration is that we only accept positive \texttt{Deposit}s, so we interpret money in \texttt{Deposit} inputs as the maximum of $0$ and the actual value specified in the \texttt{Deposit} input.

@@

A conservative upper bound for the expiration slot number can be calculated efficiently by using the function \texttt{maxTime} (or \texttt{maxTimeContract} in the Isabelle semantics), essentially by taking the lowest upper bound of all the timeouts in the contract.

We proved that this conservative upper bound works for every contract in a property stated as follows:

\begin{verbatim}
theorem timedOutTransaction_closes_contract :
"validAndPositive_state sta
 ⟹ iniSlot ≥ minSlot sta
 ⟹ iniSlot ≥ maxTimeContract cont
 ⟹ endSlot ≥ iniSlot
 ⟹ accounts sta ≠ [] ∨ cont ≠ Close
 ⟹ isClosedAndEmpty
         (computeTransaction ⦇ interval = (iniSlot, endSlot)
                             , inputs = [] ⦈ sta cont)"
\end{verbatim}

\noindent
where \texttt{isClosedAndEmpty} is defined as follows:

\begin{verbatim}
fun isClosedAndEmpty :: "TransactionOutput ⇒ bool" where
"isClosedAndEmpty (TransactionOutput txOut) =
   (txOutContract txOut = Close
      ∧ accounts (txOutState txOut) = [])" |
"isClosedAndEmpty _ = False"
\end{verbatim}
\noindent
Because the money is preserved and we know that there is an empty transaction that can be sent at any point after \texttt{maxTime} that will empty the accounts of the contract, we can conclude that money will not be locked in the contract forever.

@@

In other words, for every list of \texttt{Transaction}s, there exists at least one list of \texttt{Transaction}s with length less or equal to \texttt{maxTransactionsInitialState} that has the same effect on the contracts. 
Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.

>>>>

We can also use proof assistants to demonstrate that the Marlowe semantics presents certain desirable properties, such as that money is preserved and anything unspent is returned to users by the end of the execution of any contract.

Currently, we have translated the Haskell Marlowe semantics to Isabelle while keeping both versions as close as possible, but we decided to make them different in two main aspects:

\begin{itemize}
    \item We use \emph{integers for identifiers} because they are easier to handle than strings.
    \item We use a \emph{custom implementation of maps and sets} that use lists because Isabelle already provides many theorems that are proved for lists.
\end{itemize}

@@

Isabelle automatically proves termination for most function. This is not the case for \texttt{reductionLoop}. This function repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination requires a proof that \texttt{reduceContractStep} will eventually do that.

We proved that by establishing a measure for the size of a pair of \texttt{Contract} and \texttt{State}:

@@

We need the number of accounts in the \texttt{State} because the size of the contract \texttt{Close} will not decrease when calling \texttt{reduceContractStep}, but the number of accounts will, unless they are all empty.

And we needed to multiply the size of the \texttt{Contract} by two because the primitive \texttt{Deposit} may increase the number of accounts by one, so we need to multiply the effect of the reduction of the size of the contract in order to compensate that.

@@

There are some values for \texttt{State} that are allowed by its type but make no sense, especially in the case of Isabelle semantics where we use lists instead of maps and sets:

\begin{enumerate}
    \item The lists represent maps, so they should have no repeated keys.
    \item We want two maps that are equal to be represented the same, so we force keys to be in ascending order.
    \item For the map of accounts, we only want to record those that have positive amounts.
\end{enumerate}
\noindent
We call a value for \texttt{State} \emph{valid} if the two first properties are true. And we say it has \emph{positive accounts} if the third property is true.

We have proved that functions in the semantics preserve all three properties.

@@

A contract is \emph{quiescent} if and only if the root construct \texttt{When}, or if the contract is \texttt{Close} and all accounts are empty.

We have proved that, if an input \texttt{State} is valid and all accounts are positive, then the result from processing any transaction will always be quiescent.

@@

One of the dangers of using smart contracts is that a badly written one can potentially lock its funds forever. By the end of the contract, all the money paid to the contract must be distributed back, in some way, to a subset of the participants of the contract.
To ensure this is the case we proved two properties:

\subsubsection{Money preservation}
Money is not created or destroyed by the semantics. More specifically, the money that comes in plust the money in the contract before the transaction must be equal to the money that comes out plus the contract after the transaction, except in the case of an error, when the contract stays unaltered.

@@

A conservative upper bound for the expiration slot number can be calculated efficiently by using the function \texttt{maxTime} (or \texttt{maxTimeContract} in the Isabelle semantics), essentially by taking the maximum of all the timeouts in the contract.

We proved that this conservative upper bound is general enough for every contract, by showing that, if the contract is not closed and empty, then an empty transaction sent after \texttt{maxTime} will close the contract and leave the accounts empty.

Because of this property and \emph{Money preservation}, we can conclude that money will not be locked in the contract forever.

@@

Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.
